Explicit B-spline regularization in diffeomorphic image registration

### 2013-11-24 paper.tex
ENH:  Resubmission.

\def\journal{Neuroscience}%%% write here for which journal %%%
\def\DOI{}
\def\articleType{Methods}
\def\keyFont{\fontsize{8}{11}\helveticabold }
%\def\firstAuthorLast{Tustison {et~al.}} %use et al only if is more than 1 author
\def\firstAuthorLast{Tustison and Avants} %use et al only if is more than 1 author
\def\Authors{Nicholas J. Tustison\,$^{1,*}$ and Brian B. Avants\,$^{2}$
 }
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$University of Virginia, Department of Radiology and Medical Imaging, Charlottesville, VA, USA \\
$^{2}$Penn Image Computing and Science Laboratory, University of Pennsylvania, Department of Radiology, Philadelphia, PA, USA }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Nick Tustison}
\def\corrAddress{University of Virginia, Department of Radiology and Medical Imaging,  480 Ray C Hunt Drive, Charlottesville, VA, 22903}
\def\corrEmail{ntustison@virginia.edu}

@@

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance,
desirable properties, and several available implementations.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.  We denoted this new FFD variant as
directly manipulated free-form deformation (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

@@

As alluded to earlier, B-spline approximation can also be used for
regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein, and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to represent the
characteristic velocity fields.  However, we use the directly
manipulated free-form deformation formulation to improve the solution
convergence.  This also facilitates modeling temporal periodicity and
the enforcement of stationary boundaries.  Both this work and our
earlier work \citep{tustison2009} demonstrate that the DMFFD framework
is potentially applicable to the entire gamut of diffeomorphic
registration algorithms and provides unique ``flavors'' of smoothing
possibilities.  However, the two approaches produce characteristically
different solutions.  In addition to smoothing kernel differences,
Gaussian convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD approach.
Also, whereas Gaussian convolution operates entirely within
discretized space, the B-spline approximation routine constructs a
continuous object prior to any voxelwise reconstruction of the sampled
fields.  Interestingly, a similar comparison was made with respect to
Gaussian derivative estimation \citep{bourma2007}.  Although typically
estimated using truncated, discrete Gaussian convolution, an
alternative based on B-spline approximation demonstrated superior
performance with similar computational cost.
%Further elaboration on these differences
%are discussed in subsequent sections.

@@

Therefore, in addition to the full-scale
SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in differents applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is due to the restriction of the discrete, time-parameterized
velocity field samples to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Furthermore, the forward and inverse
mappings are guaranteed to be consistent within the discrete domain i.e. $\|
\phi_i^{-1}(\phi_i) - \mathbf{Id} \|^2 < \epsilon$.   
Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

\begin{algorithm}
\caption{Greedy SyN algorithm}
\label{alg:syn}
\begin{algorithmic}
\State $\phi_i \leftarrow \mathbf{Id}$, $\phi_i^{-1} \leftarrow \mathbf{Id}$
\Comment $i \in \{1,2\}$
\State $n \leftarrow 1$
\While {not converged}
  \State $v_1^n \leftarrow \nabla \Pi_{\sim} \left(I\circ\phi_1^{n-1},J\circ\phi_2^{n-1}\right)$
  \State $v_2^n \leftarrow \nabla \Pi_{\sim} \left(J\circ\phi_2^{n-1},I\circ\phi_1^{n-1}\right)$
  \State $v_i^n \leftarrow S_v( v_i^n )$      \Comment $S_v$ is a smoothing operation on the update transform field
  \State $\phi_i^n \leftarrow S_\phi( v_i^n \circ \phi_i^{n-1} )$      \Comment $S_\phi$ is a smoothing operation on the total transform field
  \State $\left(\phi_i^n\right)^{-1} \leftarrow Inv\left(\phi_i^n, \left(\phi_i^{n-1}\right)^{-1}\right)$
    \Comment Inverse field estimation described in \cite{avants2008}
  \State $n \leftarrow n + 1$
\EndWhile
\State \Return $\phi \leftarrow \phi_1 \circ \phi_2^{-1}$, $\phi^{-1} \leftarrow \phi_2 \circ \phi_1^{-1}$
\end{algorithmic}
\end{algorithm}

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  As mentioned earlier, 
a viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation, spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient at voxel $c$.

@@

Three diffeomorphic algorithms described previously utilizing Gaussian convolution smoothing and their DMFFD counterparts are available through the following classes in the ITK repository: 
\begin{itemize}
  \item {\bf LDDMM:}
    \begin{itemize}
      \item {\tt itk::TimeVaryingVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::TimeVaryingVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::GaussianSmoothingOnUpdateTimeVaryingVelocityFieldTransform}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf DARTEL:}
    \begin{itemize}
       \item{\tt itk::ConstantVelocityFieldTransform}
       \item{\tt itk::ConstantVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::GaussianExponentialDiffeomorphicTransform} 
      \item {\tt itk::GaussianExponentialDiffeomorphicTransformParametersAdaptor}
      \item {\tt itk::BSplineExponentialDiffeomorphicTransform} 
      \item {\tt itk::BSplineExponentialDiffeomorphicTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf SyN:}
    \begin{itemize}
      \item {\tt itk::SyNImageRegistrationMethod} 
      \item {\tt itk::BSplineSyNImageRegistrationMethod} 
    \end{itemize}  
\end{itemize}
These and other classes (e.g. similarity metrics, optimization methods, and
utility classes) were developed as part of the ITKv4 registration
framework refactoring.  Further details can be found in the documentation
provided within the classes themselves.%
\footnote{
http://www.itk.org/Doxygen/html/classes.html
}

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible). 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Any pair of images can be specified per metric per stage.

@@

In the well-known Klein comparative study \citep{klein2009}, 14 image registration
algorithms were evaluated based on performance on publicly available labeled brain data.
%Based on the final rankings, the SyN algorithm was determined to be one
%of the top-performing deformable registration algorithms.  Its relative
%performance coupled with its public availability has made
%it a de facto standard for algorithmic comparison.
For our evaluation, we used these same data.  
Specifically, we used the data sets denoted as:
\begin{itemize}
  \item MGH10 
  \item CUMC12
  \item IBSR18
  \item LPBA40
\end{itemize}
which are available for download from Arno Klein's website.%
\footnote{
http://mindboggle.info/papers/evaluation\_NeuroImage2009.php
}
The number of subjects per cohort is provided in the denotation.
Further details of these labeled brain data (e.g. labeling protocol,
data sources) are given in \citep{klein2009}.
We also include the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} 
which we denote as MAL35.  This T1-weighted
MRI data set consists of 35 subject MRIs taken from the
Oasis database.%
\footnote{
http://www.oasis-brains.org
}  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

@@

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  These templates were built using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

@@

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with ANTs.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus
labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.

>>>>

\def\journal{Neuroinformatics}%%% write here for which journal %%%
\def\DOI{}
\def\articleType{Methods}
\def\keyFont{\fontsize{8}{11}\helveticabold }
%\def\firstAuthorLast{Tustison {et~al.}} %use et al only if is more than 1 author
\def\firstAuthorLast{Tustison and Avants} %use et al only if is more than 1 author
\def\Authors{Nicholas J. Tustison\,$^{1,*}$ and Brian B. Avants\,$^{2}$
 }
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$University of Virginia, Department of Radiology and Medical Imaging, Charlottesville, VA, USA \\
$^{2}$Penn Image Computing and Science Laboratory, University of Pennsylvania, Department of Radiology, Philadelphia, PA, USA }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Nick Tustison}
\def\corrAddress{University of Virginia, Department of Radiology and Medical Imaging,  480 Ray C Hunt Drive, Charlottesville, VA, 22903}
\def\corrEmail{ntustison@virginia.edu}

@@

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance, and
{\color{blue} guaranteed parametric continuity}.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.%
\footnote{
{\color{blue}{This phenomenon is explained in greater detail in \cite{tustison2009}
but, briefly, the distribution of the uniform B-spline basis functions
spanning the image domain produces long ravines in the image registration
energy topography.  These topographical features cause an
inefficient back-and-forth traversal in classic steepest descent optimization, also known as ``hemstitching.''}}
}
We denoted this new FFD variant as
{\it directly manipulated free-form deformation} (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

@@

As alluded to earlier, B-spline approximation can also be used for
regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein, and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to represent the
characteristic velocity fields.  However, we use the directly
manipulated free-form deformation formulation to improve the solution
convergence.  This also facilitates modeling temporal periodicity and
the enforcement of stationary boundaries.  Both this work and our
earlier work \citep{tustison2009} demonstrate that the DMFFD framework
is potentially applicable to the entire gamut of diffeomorphic
registration algorithms and provides {\color{blue} alternative smoothing 
possibilities with different continuity properties (e.g. $C^2$
versus $C^3$)}.

However, the two approaches produce characteristically
different solutions.  In addition to smoothing kernel differences,
Gaussian convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD approach.
Also, whereas Gaussian convolution operates entirely within
discretized space, the B-spline approximation routine constructs a
continuous object prior to any voxelwise reconstruction of the sampled
fields.  Interestingly, a similar comparison was made with respect to
Gaussian derivative estimation \citep{bourma2007}.  Although typically
estimated using truncated, discrete Gaussian convolution, an
alternative based on B-spline approximation demonstrated superior
performance with similar computational cost.
%Further elaboration on these differences
%are discussed in subsequent sections.

@@

Therefore, in addition to the full-scale
SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in different applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is due to the restriction of the discrete, time-parameterized
velocity field samples to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Furthermore, the forward and inverse
mappings are guaranteed to be consistent within the discrete domain i.e. $\|
\phi_i^{-1}(\phi_i) - \mathbf{Id} \|^2 < \epsilon$.   
Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

\begin{algorithm}
\caption{Greedy SyN algorithm}
\label{alg:syn}
\begin{algorithmic}
\State $\phi_i \leftarrow \mathbf{Id}$, $\phi_i^{-1} \leftarrow \mathbf{Id}$
\Comment $i \in \{1,2\}$
\ForAll {image resolution levels} 
\State $n \leftarrow 1$
\While {not converged}
  \State $v_1^n \leftarrow \nabla \Pi_{\sim} \left(I\circ\phi_1^{n-1},J\circ\phi_2^{n-1}\right)$
  \State $v_2^n \leftarrow \nabla \Pi_{\sim} \left(J\circ\phi_2^{n-1},I\circ\phi_1^{n-1}\right)$
  \State $v_i^n \leftarrow S_v( v_i^n )$      \Comment $S_v$ is a smoothing operation on the update transform field
  \State $\phi_i^n \leftarrow S_\phi( v_i^n \circ \phi_i^{n-1} )$      \Comment $S_\phi$ is a smoothing operation on the total transform field
  \State $\left(\phi_i^n\right)^{-1} \leftarrow Inv\left(\phi_i^n, \left(\phi_i^{n-1}\right)^{-1}\right)$
    \Comment Inverse field estimation described in \cite{avants2008}
  \State $n \leftarrow n + 1$
\EndWhile
  \State upsample current $\phi_i$ and $\phi_i^{-1}$ to next resolution level 
  \Comment $i \in \{1,2\}$
\EndFor
\State \Return $\phi \leftarrow \phi_1 \circ \phi_2^{-1}$, $\phi^{-1} \leftarrow \phi_2 \circ \phi_1^{-1}$
\end{algorithmic}
\end{algorithm}

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  A viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.

Given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is given by
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation, spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient at voxel $c$.%
\footnote{
{\color{blue}{For comparison, FFD image registration uses the
calculated gradient which, for a single control point is evaluated using (in 
discretized form assuming no other explicit regularization):
\begin{align}
\label{eq:ffd}
  \delta v_{i_1,\ldots,i_d} = \sum_{c=1}^{N_{\Omega}} 
  \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c 
  \prod_{j=1}^d B_{i_j}(x_j^c).
\end{align}
Intuitively, this can be understood as a distribution of the spatial similarity metric
gradient to the set of control points according to the basis function weighting which, due to
the local property of B-splines, is only non-zero in a localized region surrounding each 
sample point.  The problematic issue with standard FFD image registration is that 
the parametric domain of 
the peripheral B-spline basis functions extends beyond the image boundaries resulting in
a relatively lower weighting contribution effect.  As a corrective, Equation (\ref{eq:dmffd})
normalizes the control point gradient contribution by the actual B-spline weighting overlap
with the image domain.

Note that this modification is first discussed and derived in \cite{hsu1992,lee1997} in the context of fitting scattered data for a computer graphics audience which was an extension and improvement over Sederberg's original FFD manipulation computer graphics technique \citep{sederberg1986}.  Whereas the latter (i.e. FFD) technique performs geometric deformation of the graphics object embedded in a B-spline object via manipulation of the control points, the former (i.e. DMFFD) technique permits direct manipulation of the actual
object with the control point values being updated indirectly.  The corresponding fast algorithm for scattered data approximation using DMFFD was proposed in \cite{lee1997}.

Based on the needs of one of our colleagues, the first author (N.T.) implemented and
generalized the work of \cite{lee1997}  which was reported in \cite{tustison2006} and
was N.T.'s very first ITK contribution.  
A couple years later the second author (B.A.) requested an FFD implementation for 
ANTs which N.T. naively assumed to be equivalent to applying \cite{tustison2006}
to smoothing the metric gradient (a la Demons).  The subsequent realization 
that this assumption was incorrect and figuring out why the algorithm still worked 
despite it not being a correct FFD implementation resulted in \cite{tustison2009}.
}}
}

@@

Three diffeomorphic algorithms described previously utilizing Gaussian convolution smoothing and their DMFFD counterparts are available through the following classes in the ITK repository: 
\begin{itemize}
  \item {\bf LDDMM:}
    \begin{itemize}
      \item {\tt itk::TimeVaryingVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::TimeVaryingVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::GaussianSmoothingOnUpdateTimeVaryingVelocityFieldTransform}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransform} 
      \item {\tt itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethodv4}
      \item {\tt itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf DARTEL:}
    \begin{itemize}
       \item{\tt itk::ConstantVelocityFieldTransform}
       \item{\tt itk::ConstantVelocityFieldTransformParametersAdaptor}
      \item {\tt itk::GaussianExponentialDiffeomorphicTransform} 
      \item {\tt itk::GaussianExponentialDiffeomorphicTransformParametersAdaptor}
      \item {\tt itk::BSplineExponentialDiffeomorphicTransform} 
      \item {\tt itk::BSplineExponentialDiffeomorphicTransformParametersAdaptor}
    \end{itemize}  
  \item {\bf SyN:}
    \begin{itemize}
      \item {\tt itk::SyNImageRegistrationMethod} 
      \item {\tt itk::BSplineSyNImageRegistrationMethod} 
    \end{itemize}  
\end{itemize}
These and other classes (e.g. similarity metrics, optimization methods, and
utility classes) were developed as part of the ITKv4 registration
framework refactoring.  {\color{blue}{Much of the original ITK image registration infrastructure 
was left intact including the so-called ``sparse'' transforms such as various
rigid (versor, Euclidean) and other linear transforms.  The transform classes
contributed by our group were, including those listed above, were meant to 
augment what already existed.  All the transforms listed above are derived from
the {\tt itk::DisplacementFieldTransform} class which permits specification of 
a transform described by a sampled displacement field.  The derived classes are
then modified according to the different transform constraints. Other types
of classes are used to coordinate the image registration process.  The {\tt itk::ImageRegistrationMethodv4} is the base interface for performing all 
image registration steps.  Smoothing and resampling for multi-resolution
image registration is performed in this class as is the calling of the
selected optimizer.  Output consists of a single optimized transform.
Multiple instantiations of this class in series, in conjunction with the 
{\tt itk::CompositeTransform} class, are used to optimize a composition of transforms.
Some of the diffeomorphic approaches do not easily fit into this
generalized optimization framework necessitating specialized method classes such as
those listed above.  The adaptors%
\footnote{
{\color{blue}{We gratefully acknowledge Marius Staring's contribution in providing the broad
design of the adaptors.}}
}
are used to modify the parameters between resolution levels during the course
of transform optimization within the methods classes.  For example, the 
resolution of the displacement field transforms follows that of the image 
resolution and the updating is handled by the corresponding transform adaptors.
}}
Further details can be found in the documentation
provided within the classes themselves.%
\footnote{
http://www.itk.org/Doxygen/html/classes.html
}

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible).%
\footnote{
{\color{blue}{This step corresponds to the {\tt --initial-moving-transform} option.  Generally, the
user can specify an ITK transform for initialization or can perform an initial
translation based on either the geometric center of the images, the center of the
image intensities, or the origin of the images.  See the help menu `{\tt antsRegistration --help}' for more details.
}}
} 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Any pair of images can be specified per metric per stage.%
\footnote{
{
\color{blue}{
To help the reader who wishes to explore {\tt antsRegistration} with specific use of the
parameters used for this study and specified above, we have created a script with a simplified interface and placed it in the {\tt Scripts/} subdirectory of the ANTs repository.  This script, called {\tt antsRegistrationSyN.sh} takes a fixed and moving image and performs a comprehensive (i.e. rigid $\rightarrow$ affine $\rightarrow$ deformable) registration using SyN (`{\tt -t s}') or B-spline SyN (`{\tt -t b}').
}}
}

@@

In the well-known Klein comparative study \citep{klein2009}, 14 image registration
algorithms were evaluated based on performance on publicly available labeled brain data.
%Based on the final rankings, the SyN algorithm was determined to be one
%of the top-performing deformable registration algorithms.  Its relative
%performance coupled with its public availability has made
%it a de facto standard for algorithmic comparison.
For our evaluation, we used these same data.  
Specifically, we used the data sets denoted as:
\begin{itemize}
  \item CUMC12
  \item IBSR18
  \item LPBA40
  \item MGH10 
\end{itemize}
which are available for download from Arno Klein's website.%
\footnote{
http://mindboggle.info/papers/evaluation\_NeuroImage2009.php
}

The number of subjects per cohort is provided in the denotation.
{\color{blue}{Table 1 summarizes core information about the data sets used.}}
Further details of these first four labeled brain data (e.g. labeling protocol,
data sources) are given in \citep{klein2009}.
We also include the labeled brain data provided at the
MICCAI 2012 Grand Challenge and Workshop on Multi-Atlas
Labeling%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
} 
which we denote as MAL35.  This T1-weighted
MRI data set consists of 35 subject MRIs taken from the
Oasis database.%
\footnote{
http://www.oasis-brains.org
}  The corresponding labels were provided by Neuromorphometrics,
Inc.%
\footnote{
http://Neuromorphometrics.com/
}
under academic subscription.

\begin{table*}[htb]
  \caption{
    {\color{blue}{ Brief overview of the data sets used for the SyN vs. B-spline SyN comparison.}}
    }
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c  c l}
  \toprule
  \multicolumn{1}{c}{Cohort} & Resolution & Number of labels & \multicolumn{1}{c}{Preprocessing} \\
  \midrule
  CUMC12 & $[0.86,0.86,1.5]$ & 131 & \parbox[t]{6.5cm}{rotated to ``cardinal'' pose}\\ 
  IBSR18 & $[1,1.5,1]$ & 96 & \parbox[t]{6.5cm}{aligned to Talaraich, bias corr.}\\ 
  LPBA40 & $[0.86,1.5,0.86]$ & 56  & \parbox[t]{6.5cm}{registered to MNI305, bias corr.}\\
  MGH10 & $[1,1,1]$ & 106 & \parbox[t]{6.5cm}{affine-registered to MNI152, bias corr.}\\ 
  MAL35 & $[1,1,1]$ & 138 & \parbox[t]{6.5cm}{average of 3-4 acquisitions, bias corr.}\\
  \bottomrule
  \end{tabular*}
  \label{table:demo}
\end{table*}

@@

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  {\color{blue}{Not only does this reduce the computational
time required for finding the pairwise transforms between subjects but 
prior work has demonstrated  improvement in registration with this approach\
over direct pairwise registration \citep{klein2010a}.  Since the two algorithms
have been implemented within the same framework, all registration parameters are 
identical (i.e. linear registration stage parameters, winsorizing values, etc.) 
except for the parameters governing the smoothing of the gradient field. 
}}

The cohort templates were built using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

@@

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with ANTs.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.%
\footnote{
{\color{blue}{The MALF labels are available at http://figshare.com/account/projects/196.}}
}
